Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2012-09-06 Dail_Mad_ext.tex
working on DM overview

@@

The DM model requires count data collected at $R$ sites, each of which
is surveyed on $T$ time periods, such as years. A site is ideally, but
not necessarily, a
well-defined region of a study area, such as a wetland or a patch of
early-successional habitat. In some cases, a site may not have a clear
biological significance, for instance in the case of randomly located
survey plots. The timeframe of the study is arbitrary, but the general
situation is one in which the time period is sufficiently long such
that abundance changes at each site. We will call
these time periods primary sampling periods to distinguish them from
secondary sampling periods, which are repeated surveys within a
primary period. In the case where no secondary sampling was conducted,
let $n_{it}: i=1,\hdots,R; t=1,\hdots,T$ denote the count data at site
$i$ and primary period $t$. If $J$ secondary sampling periods were used, $n_{ijt}:
j=1,\hdots,J$ is the count at site $i$ during secondary sampling
occasion $j$ within year $t$. These are the observed data, which
typically will less than the actual quantitiy of interest, abundance,
denoted $N_{it}$. In cases where detection probability is perfect,
abundance is observed directly such that $\{N_{it}\}$ are the data.

@@

As originally formulated, the DM model includes submodels for four
conditionally related processes: the initial abundance state, apparent
survival (the number of individuals that survive and do not
emmigrate), recruitment (the number of new individuals entering
the population), and detection probability \citet{dail_madsen:2011}.
Taken together, the first 3 processes describe the state process---the
variation in abundance in space and time. The forth item in the list
is the observation process, which describes the relationship between
the true state, abundance, and the observed count data.

@@

In the original formulation,
initial abundance---abundance at site $i$ during primary period 1, was
modeled using either the Poisson or negative binomial distributions:
\begin{gather}
N_{i1} \sim \mathrm{Pois}(\Lambda_i) \nonumber \\
N_{i1} \sim \mathrm{NB}(\Lambda_i, \alpha)
\label{eq:N1}
\end{gather}
where
%$N_{i1}$ is the abundance at site $i=1,\hdots,R$ during year 1 and
$\Lambda_i$ is the expected abundance at site $i$ during year 1. The
Poisson distribution assumes that the mean of $N_{i1}$ is equal to its
variance---an assumption relaxed when using the negative binomial
distribution, which includes the overdispersion parameter $\alpha$.

In either case, the model for initial abundance has two distinguising
features. First, it allows one to model spatial variation in
abundance. For instance, one might consider the influence of some
environmental covariate on abundance using a simple generalized linear
model such as $\log(\Lambda_i) = \beta^{\Lambda}_0 + \beta^{\Lambda}_1
x_{i}$ which is a log-linear function with an intercept and slope
parameter for the effect of $x_i$ on the logarithm of $\Lambda$.

The second thing to note is that the spatial replicates resolve the
problem of parameter non-identifiability that are common to standard
state-space models.

\subsubsection{Abundance after $t=1$}

The survival and recruitment terms allow abundance to change over time:
\begin{equation}
\left.\begin{aligned}
S_{it}|N_{it-1} &\sim \mathrm{Bin}(N_{it-1}, \omega) \\
G_{it}|N_{it-1} &\sim \mathrm{Pois}(\gamma(N_{it-1})) \\
N_{it} &= S_{it}+G_{it}
\end{aligned}\right\} \quad \text{for} \; t=2,\hdots,T
\label{eq:Nt}
\end{equation}
where $\omega$ is the survival probability and $\gamma$ is the arrival
rate (which can depend on $N_{it-1}$). Dail and Madsen propose three models
for $\gamma$: the constant model, where $\gamma$ does not depend on
$N_{it-1}$, and which simulates a ``propagule rain'' of recruitment; the
autoregressive model , which
simulates geometric or density independent growth; and the ``no-trend''
model, which keeps abundance fixed over time.

Eqs.~\ref{eq:N1} and \ref{eq:Nt} fully specify the state model---the
model for spatial and temporal variation in abundance. The observation
model assumes that individuals are missed due to imperfect
detection. The simplest model for imperfect detection is
\begin{equation}
  n_{it} \sim \mathrm{Bin}(N_{it}, p)
  \label{eq:p1}
\end{equation}

Initial abundance, apparent survival, recruitment, and detection can
all be modeled as functions of covariates, allowing fairly
sophisticated inference about population processes and dynamics from
point count data.

@@

\subsection{Range model of initial abundance}

@@

@@

\subsection{Dynamic range models}

>>>>

The DM model is an extension of the $N$-mixture model
\citep{royle:2004biom}, which allows for inference about spatial
variation in abundance when individuals cannot be detected with
certainty. To estimate both abundance parameters and parameters of the
detection process, the model uses replicate observations at each site,
which are collected during sufficiently short time intervals such that
the population can safely be assumed to be closed with respect to
births, deaths, and movement. The DM model relaxes this closure
assumption and includes explicit parameters
describing population change over time.

@@

The DM model requires count data collected at $R$ sites, each of which
is surveyed on $T$ time periods. A site is ideally, but not necessarily, a
well-defined region of a study area, such as a wetland or a patch of
early-successional habitat. In some cases, a site may not have a clear
biological significance, for instance in the case of randomly located
survey plots. The timeframe of the study is arbitrary, but the general
situation is one in which the time period is sufficiently long such
that abundance changes at each site. We will call
these time periods primary sampling periods to distinguish them from
secondary sampling periods, which are repeated surveys within a
primary period. In the case where no secondary sampling was conducted,
let $n_{it}: i=1,\hdots,R; t=1,\hdots,T$ denote the count data at site
$i$ and primary period $t$. If $J$ secondary sampling periods were used, $n_{ijt}:
j=1,\hdots,J$ is the count at site $i$ during secondary sampling
occasion $j$ within year $t$. These are the observed data, which
typically will less than the actual quantitiy of interest, abundance,
denoted $N_{it}$. In cases where detection probability is perfect,
abundance is observed directly such that $\{N_{it}\}$ are the data.

@@

As originally formulated, the DM model includes submodels for three
conditionally related processes: (1) initial abundance, i.e. the
abundance at site $i$ during the first primary period,
denoted $[N_{i1}]$, (2) subsequent abundance, which is dependent upon
abundance in previous years $[N_{it}|N_{it-1}]$, and (3) the
detection process, $[n_{it}|N_{it}]$ \citep{dail_madsen:2011}.
The first two processes describe the state process---the
variation in abundance in space and time. The third item in the list
is the observation process, which describes the relationship between
the true state, abundance, and the observed count data.

@@

In the original formulation, initial abundance was
modeled using either the Poisson or negative binomial distributions:
\begin{gather}
N_{i1} \sim \mathrm{Pois}(\Lambda_i) \nonumber \\
\text{or} \nonumber \\
N_{i1} \sim \mathrm{NB}(\Lambda_i, \alpha)
\label{eq:N1}
\end{gather}
where
%$N_{i1}$ is the abundance at site $i=1,\hdots,R$ during year 1 and
$\Lambda_i$ is the expected abundance at site $i$ during year 1. The
Poisson distribution assumes that the mean of $N_{i1}$ is equal to its
variance, whereas the negative binomial distribution allows the
variance to be greater than the mean with the amount of overdispersion
determined by the parameter $\alpha$.

Regardless of the specified distribution, the model for initial
abundance has two distinguising features. First, it provides a
mechanism for characterizing spatial variation in abundance. For
instance, one might consider the influence of some environmental
covariate ($x_i$) on abundance using a log-linear
model such as $\log(\Lambda_i) = \beta^{\Lambda}_0 + \beta^{\Lambda}_1
x_{i}$. The second important point is that the spatial replicates resolve the
problem of parameter non-identifiability that are common to standard
state-space models because, as demonstrated by \citet{royle:2004biom},
process variation and observation error can be estimated from
spatially-replicated count data. Hence, the first component of the
model addresses issues (1) and (2) identified in the introduction.

\subsubsection{Abundance in subsequent time periods}

The DM model assumes that abundance in time $t$ is a function of
abundance in time $t-1$, i.e. abundance at each site evolves as a
first order Markovian process, although
higher order processes are also possible.  \citet{dail_madsen:2011}
considered several models to describe the temporal dynamics; however,
in each case they modeled $N_t$ as the sum of two random variables:
$S_t$, the number of indivduals surviving from $t-1$ and not
emigrating; and $G_t$ the number of new individuals entering the
population. Their most general model was
\begin{equation}
\left.\begin{aligned}
S_{it}|N_{it-1} &\sim \mathrm{Bin}(N_{it-1}, \omega) \\
G_{it}|N_{it-1} &\sim \mathrm{Pois}(\gamma(N_{it-1})) \\
N_{it} &= S_{it}+G_{it}
\end{aligned}\right\} \quad \text{for} \; t=2,\hdots,T
\label{eq:Nt}
\end{equation}
where $\omega$ is the apparent survival probability and $\gamma$ is
the recruitment rate (which can depend on $N_{it-1}$). In the absence
of movement, $\omega$ is exactly the probability of surviving from
year $t$ to $t-1$, and $\gamma$ is the per-capita birth rate. However,
in the more common scenario when immigration and emigration occur,
$\omega$ is the probability of surviving and not emigrating, and
$\gamma$ is product of the birth rate and the emigration rate,
i.e. the recruitment rate. \citet{dail_madsen:2011} proposed three
models for $\gamma$: the constant model,
$G_t \sim \mathrm{Pois}(\gamma)$ where recruitment does not depend on
$N_{it-1}$, and which simulates a ``propagule rain'' of new individuals; the
autoregressive model, $G_t \sim \mathrm{Pois}(\gamma(N_{it-1}))$, which
simulates geometric or density independent growth; and the
``no-trend'' model, $\gamma = (1-\omega)\Lambda$, which keeps
expected abundance constant over time. As before, covariates of
$\omega$ and $\gamma$ can be easily accomodated, for example using
logit- and log-linear models respectively.

In cases where no movement occurs, the DM model can directly estimate
demographich parameters, and this mechanistic description of
population change is much more appealing than classical
population growth models such as the logistic model. However, movement
complicates the interpretation of the the parameters, reducing the
connection to specific population processes. In such cases, the model
of dynamics shown in Eq~\ref{eq:Nt} can be replaced with standard
population models as described in Sect~\ref{sec:ext}. Irrespective of
the model for population dynamics, it is worthwhile to note that the
discrete distributions used in the DM model avoid the issues
associated with the Gaussian distribution used to model process
variation in standard state-space models.

\subsubsection{Observation process}

Eqs.~\ref{eq:N1} and \ref{eq:Nt} fully specify the state model---the
model for spatial and temporal variation in abundance. The observation
model assumes that individuals are missed due to imperfect
detection. The simplest model for imperfect detection is
\begin{equation}
  n_{it} \sim \mathrm{Bin}(N_{it}, p)
  \label{eq:p1}
\end{equation}
Surprsingly detection probability, $p$, can be estimated without
secondary sampling occasions. However, this is only possible because
if the parameteric assumptions of the population dynamics are met
exactly. In practice, it is preferable to obtain direct information
about $p$ using secondary sampling occassions or some other auxiliary
data, as is demonstrated in section

@@

@@

%\subsection{Range model of initial abundance}
\subsection{Excess zeros}

%\subsection{Dynamic range models}

@@

\subsection{Bayesian inference}

priors are good
